% ============================================================================
%   ABSTRACT (English) — Aivancity: 250-300 words, ≥5 keywords, self-contained
% ============================================================================

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\markboth{Abstract}{}

Patient matching to clinical trials is a bottleneck in translational 
medicine, and recent large language model (LLM) systems such as TrialGPT 
aim to automate the assessment of patient eligibility against inclusion 
and exclusion criteria. Beyond raw accuracy, however, the probabilistic 
reliability of these systems has not been systematically audited. A 
recent scoping review of twenty-four LLM-based matching systems 
identified zero that report calibration metrics. This omission is 
consequential: an overconfident model that misclassifies with high 
confidence is clinically worse than an accurate but honestly uncertain 
one.

We conducted the first systematic calibration audit of open-source LLMs 
in this domain. Twelve configurations (six models crossed with two 
prompt variants) were evaluated on the 1{,}015 patient--criterion pairs 
of TrialGPT-Criterion-Annotations. Four pre-registered hypotheses were 
tested with permutation tests and Bonferroni correction, and stability 
was verified through five-fold cross-validation and comparison to 
naive baselines. Probability distributions were extracted through 
direct log-probability scoring rather than free-form generation.

All twelve configurations exhibit Expected Calibration Error above the 
clinical threshold of $0.05$, with values ranging from $0.124$ to 
$0.602$. Reinforcement learning from human feedback significantly 
degrades calibration on both the Mistral-7B and Llama-3.1-8B families 
(all four tests, $p < 0.003$). Three novel findings emerge from deeper 
analysis: enriched expert prompts degrade calibration in five of six 
models; five configurations exhibit mode collapse, predicting a single 
label in more than $97\%$ of cases; on the $238$ pairs requiring 
truly discriminative decisions, both medical LLMs tested (Meditron and 
BioMistral) achieve accuracy at or below $0.4\%$. Post-hoc isotonic 
recalibration brings eleven of twelve configurations below the 
calibration threshold but cannot restore reasoning to collapsed models.

Aggregate calibration and accuracy metrics can mask complete failure of 
open-source medical LLMs on clinically decisive cases. A responsible 
evaluation protocol must combine calibration, prediction diversity, and 
stratified analysis by case difficulty.

\vspace{0.5cm}
\noindent\textbf{Keywords:} large language models, calibration, 
clinical trial matching, RLHF, medical AI, mode collapse, expected 
calibration error.

\cleardoublepage