% ============================================================================
%   CHAPTER 2 — STATE OF THE ART
%   Aivancity: 10-15 pages, thematic organization, ≥15 peer-reviewed refs
% ============================================================================

\chapter{State of the Art}
\label{chap:sota}

This chapter reviews the state of the art at the intersection of three 
research areas that jointly frame the present study: patient--clinical 
trial matching, probabilistic calibration of classifiers, and the 
evaluation of medically specialized large language models. Rather than 
surveying each area in isolation, we organize the review around the 
specific gap that motivates this thesis --- the absence of systematic 
calibration audits in LLM-based clinical trial matching --- and use 
each section to build the position of this work with respect to the 
closest prior contributions. Section~\ref{sec:sota-positioning} closes 
the chapter with a comparative table and an explicit statement of the 
contribution.

% ----------------------------------------------------------------------------
\section{Patient--Clinical Trial Matching}
\label{sec:sota-matching}

\subsection{Clinical Context and Recruitment Bottlenecks}
\label{subsec:sota-clinical}

Clinical trial recruitment is one of the most persistent bottlenecks in 
translational medicine. Despite widespread interest from patients and 
physicians, less than 5\% of adult cancer patients enroll in a clinical 
trial, and 80\% of trials fail to meet their recruitment 
targets~\cite{jin2024trialgpt}. The mismatch between patients and 
appropriate trials is caused, in part, by the sheer volume of open 
studies (over 400{,}000 registered on ClinicalTrials.gov at the time of 
writing) and the complexity of their eligibility criteria, which 
combine numeric thresholds, medical history requirements, and drug 
exclusions in dozens of criteria per trial. Manual screening of a 
patient against this space of criteria is labor-intensive and 
error-prone, and automating even a fraction of this process would 
substantially accelerate recruitment.

Automated matching systems have thus been a topic of active research 
for over a decade. The field has passed through three distinct waves, 
each corresponding to a different technological substrate.

\subsection{Pre-LLM Approaches: Rule-Based and Embedding Systems}
\label{subsec:sota-prelm}

The first wave of matching systems relied on structured information 
extraction. EliXR~\cite{weng2011elixr}, EliIE~\cite{kang2017eliie}, and 
Criteria2Query~\cite{yuan2019criteria2query} transformed free-text 
eligibility criteria into structured OMOP or SQL queries, matched 
against structured electronic health records. These systems produce 
deterministic Boolean outputs (match or no match) and provide no 
probability estimates or uncertainty quantification. They perform well 
on criteria that map cleanly to structured concepts (age, diagnoses, 
laboratory values) but struggle with criteria requiring nuanced 
interpretation.

The second wave introduced end-to-end neural matching. 
DeepEnroll~\cite{zhang2020deepenroll} encoded criteria with BERT and 
patient EHRs with hierarchical embeddings, framing matching as textual 
entailment and reporting PR-AUC and F1 scores. 
COMPOSE~\cite{gao2020compose} added a cross-modal pseudo-siamese 
network with a composite inclusion/exclusion loss, achieving 98.0\% 
patient--criterion AUC and 83.7\% patient--trial accuracy. Neither 
system reported calibration metrics. This omission is not accidental: 
the primary evaluation targets of these systems were discrimination 
metrics (AUC, F1) rather than the reliability of the probabilistic 
outputs.

\subsection{The LLM Wave: 2023--2026}
\label{subsec:sota-llmwave}

The third and current wave dates from the release of GPT-4 and the 
subsequent proliferation of open-source large language models. The 
most influential system in this wave is 
TrialGPT~\cite{jin2024trialgpt}, published in \emph{Nature 
Communications} in 2024. TrialGPT is a three-module pipeline: a 
retrieval stage recovers more than 90\% of relevant trials in less 
than 6\% of the collection using a hybrid BM25 and MedCPT retriever; a 
matching stage emits a per-criterion ordinal label 
(``included / not included / no relevant information''); and a ranking 
stage aggregates criterion labels into trial-level scores. TrialGPT 
reports a criterion-level accuracy of 87.3\% and correlation patterns 
against the ground truth, but does not report any calibration 
metric --- no ECE, no Brier score, no reliability diagram, and no 
monotonicity check across the ordinal scale.

TrialGPT is not an isolated case. RECTIFIER~\cite{unlu2024rectifier}, 
published in \emph{NEJM AI}, applies a retrieval-augmented GPT-4 to 
the COPILOT-HF heart-failure trial and reports 97.9--100\% accuracy 
with Matthews correlation coefficients between 0.837 and 1.00. The 
system has been deployed clinically at Mass General Brigham, yet its 
outputs are reported only as accuracy and MCC, without probabilistic 
reliability assessment. PRISM and its variant 
OncoLLM~\cite{gupta2024oncollm}, published in \emph{npj Digital 
Medicine}, introduce a weighted-tier scoring method with concept-level 
importance and report top-$k$ accuracy and nDCG (68\% vs.\ 62.6\% for 
GPT-3.5) but again omit calibration. Trial-LLaMA distills open-source 
Llama on synthetic criterion labels~\cite{nievas2024triallama} and 
reports accuracy alone. Zero-shot matching approaches by Wornow et 
al.~\cite{wornow2024zeroshot} achieve 0.93 micro-F1 on the n2c2 2018 
dataset and elicit a self-declared confidence per criterion --- the 
closest any published system comes to uncertainty quantification --- 
but the self-declared confidence is never calibrated against ground 
truth. More recent systems including 
TrialMatchAI~\cite{abdallah2026trialmatchai}, multimodal pipelines by 
Callies et al.~\cite{callies2025multimodal}, and cross-modal trial-site 
selectors extend the scope of matching but continue the pattern of 
accuracy-only reporting.

The definitive scoping review of the field, published by Chen et 
al.~\cite{chen2025scoping} in \emph{JCO Clinical Cancer Informatics}, 
screened 2{,}357 papers and included 24 LLM-based matching systems, 
of which 21 were published in 2024. The review confirms that 
\emph{none} of these systems reports calibration metrics: no ECE, no 
Brier score, no reliability diagram, no calibration-in-the-large, and 
no slope. This gap defines the primary methodological opportunity 
addressed by the present thesis.

% ----------------------------------------------------------------------------
\section{Probabilistic Calibration of Classifiers}
\label{sec:sota-calibration}

\subsection{Foundations and the Modern Miscalibration Problem}
\label{subsec:sota-foundations}

Probability calibration --- the requirement that predicted 
probabilities match empirical accuracies --- has a long history in 
statistical prediction. Historical foundations include Platt 
scaling~\cite{platt1999} for logistic regression outputs, isotonic 
regression~\cite{zadrozny2002isotonic} for non-parametric monotone 
calibration, and the systematic comparisons of Niculescu-Mizil and 
Caruana~\cite{niculescu2005}. For decades, calibration was treated as 
a solved problem: shallow classifiers such as logistic regression, 
naïve Bayes, and calibrated support vector machines were well 
calibrated after appropriate post-hoc correction.

This picture changed with the rise of deep learning. Guo, Pleiss, Sun, 
and Weinberger~\cite{guo2017calibration} showed in 2017 that modern 
deep neural networks, despite their superior accuracy, are 
\emph{systematically overconfident} relative to shallower architectures: 
a network with 85\% accuracy may assign 99\% confidence to its 
predictions, including its errors. This paper introduced the Expected 
Calibration Error (ECE) as the standard metric for measuring 
miscalibration in modern classifiers, and demonstrated the 
effectiveness of temperature scaling, a single-parameter post-hoc 
method that does not modify argmax predictions.

Since 2017, the calibration literature has expanded rapidly. 
Alternative metrics have been proposed to address specific limitations 
of ECE: adaptive-ECE, class-wise ECE, and the more refined estimators 
of Nixon et al.~\cite{nixon2019measuring} and Kumar, Liang, and 
Ma~\cite{kumar2019verified}. The Brier score~\cite{brier1950} provides 
a joint measure of calibration and sharpness. Beyond ECE, 
distribution-free confidence intervals for calibration have been 
studied by Gupta and Ramdas~\cite{gupta2021distributionfree}.

\subsection{Clinical Importance of Calibration}
\label{subsec:sota-clinical-calib}

The clinical importance of calibration has been argued particularly 
forcefully by Van Calster et al.~\cite{vancalster2019calibration}, 
whose 2019 \emph{BMC Medicine} article titled ``Calibration: the 
Achilles heel of predictive analytics'' establishes calibration as a 
prerequisite for clinical deployment of any predictive model. Van 
Calster and colleagues distinguish four progressive levels of 
calibration --- calibration in the large, weak calibration, moderate 
calibration, and strong calibration --- and argue that most published 
clinical prediction models fail even the weakest level. They also 
introduce the threshold ECE $< 0.05$ as a practical benchmark for 
clinically usable calibration, which the present thesis adopts as the 
rejection criterion for Hypothesis H1 (Section~\ref{sec:met-hypotheses}).

The 2024 update of the TRIPOD guidelines, known as 
TRIPOD+AI~\cite{collins2024tripod}, formalizes calibration as a 
required reporting element for any AI-based clinical prediction model. 
Under TRIPOD+AI, a paper introducing a clinical AI model must report 
calibration curves, calibration-in-the-large, and calibration slopes; 
failure to do so is now grounds for methodological rejection by 
several leading medical journals. This regulatory shift makes the 
absence of calibration reporting in LLM-based trial matching systems 
increasingly untenable.

\subsection{Recalibration Methods}
\label{subsec:sota-recalibration}

Post-hoc recalibration methods apply a monotone transformation to 
uncorrected confidences after the classifier has been trained. The 
three methods evaluated in the present thesis --- temperature scaling, 
isotonic regression, and Platt scaling --- represent the classical 
choices; more elaborate methods include beta 
calibration~\cite{kull2017beta}, Dirichlet 
calibration~\cite{kull2019dirichlet}, and the matrix/vector scaling 
family. Intrinsic methods, which modify the training objective itself, 
include label smoothing~\cite{muller2019labelsmoothing} and focal 
loss~\cite{mukhoti2020focal}. Distribution-free calibration approaches 
combine post-hoc scaling with conformal 
guarantees~\cite{gupta2021distributionfree}.

Complementary to point-estimate calibration, split conformal 
prediction~\cite{vovk2005, angelopoulos2023conformal} produces 
prediction sets with a guaranteed marginal coverage of $1 - \alpha$ 
without any distributional assumption on the underlying model. The 
guarantees hold under exchangeability of calibration and test data, 
which is satisfied in the present study by the split protocol 
described in Section~\ref{subsec:met-data-split}. Selective 
classification --- the option to abstain from prediction --- has been 
formalized by Geifman and 
El-Yaniv~\cite{geifman2017selective, geifman2019selectivenet} through 
risk--coverage curves, which are used in Section~\ref{sec:res-h3h4} to 
test Hypothesis H4.

% ----------------------------------------------------------------------------
\section{Large Language Models: Alignment and Calibration}
\label{sec:sota-llm-calib}

The calibration of large language models raises questions that differ 
qualitatively from those of traditional classifiers. LLMs are not trained 
end-to-end for classification but for next-token prediction, and their 
class probabilities in downstream tasks are extracted through prompting 
strategies rather than through a dedicated output layer. This structural 
difference has motivated a dedicated research thread on the calibration 
of LLMs and, in particular, on the effect of post-training alignment 
procedures.

\subsection{RLHF and Its Effect on Calibration}
\label{subsec:sota-rlhf}

Reinforcement learning from human feedback (RLHF), introduced at scale 
by Ouyang et al.~\cite{ouyang2022training} in the development of 
InstructGPT, has become the dominant paradigm for aligning large 
language models with human preferences. The procedure consists of three 
stages: supervised fine-tuning on demonstration data, training of a 
reward model from human preference judgments, and reinforcement 
learning of the base model to maximize the reward. RLHF has 
demonstrably improved user-facing metrics such as helpfulness, 
truthfulness, and adherence to safety guidelines.

The calibration cost of this alignment procedure was first documented 
by Kadavath et al.~\cite{kadavath2022know} in a systematic study of 
Anthropic's Claude family. Using general question-answering benchmarks 
(TriviaQA, Lambada, TruthfulQA), the authors showed that base models 
produce well-calibrated confidence estimates on multiple-choice 
questions, while their RLHF-aligned counterparts systematically shift 
probability mass toward the model's top prediction, degrading 
calibration. Kadavath and colleagues attribute this effect to the 
alignment reward signal, which incentivizes confident, assertive 
answers regardless of their actual reliability --- because human 
raters, on average, prefer confident responses.

This finding has been replicated in several follow-up studies and is 
now considered a robust property of RLHF-based alignment. However, no 
prior work has systematically extended this observation to the clinical 
domain, where the trade-off between assertiveness and reliability 
carries substantially higher stakes. The central pre-registered 
contribution of the present thesis (Hypothesis H2, tested in 
Section~\ref{sec:res-h2}) is precisely this extension: we replicate 
Kadavath's finding on two independent open-source model families 
(Mistral-7B and Llama-3.1-8B) applied to clinical trial matching, and 
show that the effect not only persists but is quantitatively amplified 
in this domain.

\subsection{Uncertainty Elicitation and Verbalized Confidence}
\label{subsec:sota-verbalized}

An alternative to log-probability-based calibration is verbalized 
confidence elicitation, where the model is prompted to explicitly 
report its confidence in a natural-language format (e.g., ``I am 80\% 
sure that...''). This approach has been studied in general LLMs and 
was applied in the clinical domain by Wornow et 
al.~\cite{wornow2024zeroshot}, whose zero-shot clinical trial matching 
system asks GPT-4 to output a confidence score alongside each criterion 
label. Verbalized confidence is attractive for its interpretability, 
but it introduces two complications: the elicited scores are not 
necessarily connected to the model's internal probability estimates, 
and they are typically not calibrated against ground truth. The 
present thesis avoids verbalized confidence in favor of direct 
log-probability scoring (Section~\ref{sec:met-scoring}), which produces 
proper probability distributions suitable for standard calibration 
analysis.

% ----------------------------------------------------------------------------
\section{Medical Large Language Models}
\label{sec:sota-medllms}

The rise of general-purpose LLMs has been paralleled by a growing 
ecosystem of medically specialized models. These models are typically 
built by continued pretraining, instruction tuning, or model merging 
on top of a general-purpose base, using biomedical corpora such as 
PubMed abstracts, clinical guidelines, or medical textbooks. This 
section reviews the four open-source medical LLMs most relevant to 
the present study.

\subsection{Meditron}
\label{subsec:sota-meditron}

Meditron~\cite{chen2023meditron}, released by EPFL in late 2023, 
represents the first open-source medical LLM to be developed at scale. 
The 7B and 70B versions are built by continued pretraining of Llama-2 
on the GAP-Replay corpus, which consists of 48{,}000 clinical practice 
guidelines and a curated subset of PubMed abstracts totaling 
approximately 46 billion tokens. Meditron reports strong performance 
on established medical QA benchmarks (MedQA, PubMedQA, MedMCQA) but 
does not report calibration metrics. Notably, no official 
instruction-tuned or RLHF-aligned version of Meditron-7B has been 
released, which makes it the ideal candidate to isolate the effect of 
medical pretraining without any confound from post-training alignment 
(Section~\ref{sec:met-models}).

\subsection{BioMistral}
\label{subsec:sota-biomistral}

BioMistral~\cite{labrak2024biomistral}, released in early 2024, takes 
a different construction path: it is built by continued pretraining 
of Mistral-7B-Instruct-v0.1 on the PubMed Central Open Access subset 
(3 billion tokens), and then combined with the original Mistral through 
several merging strategies (DARE, TIES, SLERP). BioMistral is the only 
open-source medical LLM to date whose original publication reports ECE 
values on medical QA tasks, providing an external validation point for 
calibration measurement methodology. Because BioMistral inherits its 
alignment from Mistral-Instruct, it represents the ``medical + RLHF'' 
corner of the two-by-two matrix used in the present study.

\subsection{Med42, HuatuoGPT-o1, and the Broader Ecosystem}
\label{subsec:sota-others}

The medical LLM ecosystem has expanded rapidly since 2024. 
Med42-v2~\cite{christophe2024med42}, released by M42 Health, is 
available in 8B and 70B parameter versions and uses a full 
supervised-fine-tuning-plus-DPO pipeline on top of Llama-3. However, 
the base variant of Med42-v2 is not publicly released, which precludes 
a Base-versus-Instruct comparison of the form required for testing 
Hypothesis H2. HuatuoGPT-o1~\cite{chen2024huatuogpt}, released in late 
2024, extends the reasoning capabilities of medical LLMs through 
verification-based multi-step training, but its primary training 
corpus is Chinese, which introduces a language confound with the 
English-only benchmark used here. Other medical LLMs considered but 
excluded from the final selection include OpenBioLLM (redundant 
architecture with BioMistral), Aloe-70B (memory budget), and various 
proprietary systems (GPT-4-Medical, Med-PaLM) for which the underlying 
weights are not available for calibration audits.

The comparative characteristics of the six models selected for the 
present study are summarized in Table~\ref{tab:met-models} of 
Chapter~\ref{chap:methodology}. Two properties are shared across all 
current open-source medical LLMs: they are evaluated primarily on 
multiple-choice medical QA benchmarks, and calibration is not reported. 
The present thesis contributes to filling this gap by applying 
standard calibration protocols to four representative medical and 
generalist configurations on a common downstream task.

% ----------------------------------------------------------------------------
\section{Positioning of This Work}
\label{sec:sota-positioning}

The three research threads reviewed above --- clinical trial matching, 
probabilistic calibration, and medical LLMs --- have developed largely 
in parallel. Table~\ref{tab:sota-comparison} places the present thesis 
in the intersection of these threads, alongside the closest prior works.

\begin{table}[htbp]
    \centering
    \caption{Comparative positioning of the closest prior works. 
        A checkmark indicates that the corresponding aspect is 
        systematically addressed; a dash indicates that it is not.}
    \label{tab:sota-comparison}
    \small
    \begin{tabular}{@{}p{3cm}p{2.2cm}p{2cm}ccc@{}}
        \toprule
        \textbf{Work} & 
        \textbf{Domain} & 
        \textbf{Models} & 
        \textbf{RLHF} & 
        \textbf{Calib.} & 
        \textbf{Recalib.} \\
        \midrule
        Jin et al.\ 2024 (TrialGPT) & Trial matching & GPT-4 & --- & --- & --- \\
        Kadavath et al.\ 2022 & General QA & Anthropic & \checkmark & \checkmark & --- \\
        Labrak et al.\ 2024 (BioMistral) & Medical QA & BioMistral & --- & \checkmark & --- \\
        Unlu et al.\ 2024 (RECTIFIER) & Trial matching & GPT-4 & --- & --- & --- \\
        Nievas et al.\ 2024 (Trial-LLaMA) & Trial matching & Llama & --- & --- & --- \\
        Chen et al.\ 2025 (scoping review) & Trial matching & 24 LLMs & --- & --- & --- \\
        \midrule
        \textbf{This work} & 
        \textbf{Trial matching} & 
        \textbf{6 LLMs} & 
        \textbf{\checkmark} & 
        \textbf{\checkmark} & 
        \textbf{\checkmark} \\
        \bottomrule
    \end{tabular}
\end{table}

Three specific gaps in the literature are addressed by this thesis. 
First, no prior LLM-based clinical trial matching system reports 
calibration metrics, as documented by the Chen et 
al.~\cite{chen2025scoping} scoping review across 24 published systems. 
The present study is, to our knowledge, the first to conduct a 
systematic calibration audit in this domain. Second, no prior work has 
tested the RLHF-degrades-calibration hypothesis of Kadavath et 
al.~\cite{kadavath2022know} outside general-purpose question answering. 
The present study extends this finding to the clinical domain across 
two independent model families. Third, the standard evaluation 
protocol for medical LLMs --- multiple-choice medical QA --- masks two 
failure modes that are exposed here: mode collapse in evasion labels 
and complete failure on truly discriminative cases. The methodology 
and findings of this thesis are intended to make these failure modes 
detectable in future medical LLM evaluations.

The specific contributions of this work, summarized as four 
pre-registered hypotheses and nine original findings, are presented in 
Chapter~\ref{chap:results} and synthesized in 
Chapter~\ref{chap:conclusion}.