% ============================================================================
%   CHAPTER 6 — CONCLUSION AND FUTURE WORK
%   Aivancity: 3-5 pages
% ============================================================================

\chapter{Conclusion and Future Work}
\label{chap:conclusion}

% ----------------------------------------------------------------------------
\section{Summary of Findings}
\label{sec:conc-summary}

This thesis reports a systematic calibration audit of twelve open-source 
large language model configurations applied to patient--clinical trial 
matching. The study was structured around four pre-registered hypotheses, 
tested with fixed rejection thresholds and Bonferroni-corrected 
statistical procedures, and complemented by nine original findings that 
emerged from the deeper analysis. Table~\ref{tab:res-summary-hypotheses} 
in Chapter~\ref{chap:results} summarizes the pre-registered outcomes 
and Table~\ref{tab:res-summary-findings} lists the nine findings.

The pre-registered results support the central hypothesis of the study 
and challenge one of its ancillary claims. Hypothesis~H1 is confirmed 
across all twelve configurations: every LLM in the experimental matrix 
exhibits an Expected Calibration Error above the well-calibrated 
threshold of $0.05$~\cite{vancalster2019calibration}, with values 
ranging from $0.124$ to $0.602$. Hypothesis~H2 is confirmed on both 
model families tested (Mistral-7B and Llama-3.1-8B) and both prompt 
variants: RLHF-aligned Instruct versions exhibit significantly worse 
calibration than their Base counterparts, all four permutation tests 
rejecting the null hypothesis at Bonferroni-corrected significance 
($p < 0.003$). This is the central pre-registered contribution of the 
thesis, and it extends to the clinical trial matching domain a finding 
first documented by Kadavath et al.~\cite{kadavath2022know} in 
general-purpose question answering. Hypothesis~H3 is not supported: no 
configuration reaches the C-index threshold of $0.70$, though Meditron 
approaches it at $0.654$. Hypothesis~H4 is partially supported, with 
the selective-classification gain reaching the $+0.02$ threshold only 
for Mistral configurations.

Beyond the pre-registered hypotheses, three findings deserve particular 
emphasis. First, the enriched-prompt paradox (F3) shows that expert 
prompt engineering degrades calibration in five of six models, 
contradicting the standard recommendation to invest in prompt design as 
a route to better model behavior. Second, mode collapse (F7 and F8) 
affects five of the twelve configurations, which concentrate more than 
$97\%$ of their predictions on a single label; the collapse persists 
under constrained decoding, indicating a structural absence of clinical 
reasoning rather than a strategic use of evasion labels. Third, and 
most consequentially, on the $238$ pairs requiring a truly 
discriminative clinical decision, both medical LLMs tested (Meditron 
and BioMistral) achieve accuracy at or below $0.4\%$ (F9), despite 
their apparent superiority on global metrics. This last finding 
reverses the ranking suggested by aggregate accuracy and reveals that 
current medical LLM benchmarks systematically mask a class of failure 
that carries the highest clinical importance.

% ----------------------------------------------------------------------------
\section{Answer to the Research Question}
\label{sec:conc-answer}

The research question posed in Chapter~\ref{chap:introduction} asked 
whether the eligibility scores produced by LLM-based patient--trial 
matching systems are probabilistically reliable, and whether RLHF 
post-training affects this reliability. The answer, informed by the 
twelve configurations audited in this study, is negative on both 
dimensions.

The scores are not reliable: the systematic miscalibration documented 
by Hypothesis~H1 disqualifies every configuration from direct 
deployment as an autonomous eligibility classifier. Furthermore, the 
mode collapse behaviors documented by Findings F7 and F8 disqualify 
five of the twelve configurations from any useful classification role, 
independently of their calibration properties. The remaining seven 
configurations produce diverse predictions but remain far above the 
Van Calster threshold. Only Mistral-7B-Instruct with the minimal 
Prompt~A reaches clinically acceptable performance on the truly 
discriminative subset.

RLHF post-training significantly worsens calibration on both model 
families tested, converting the underconfidence of Mistral-Base into 
the overconfidence of Mistral-Instruct, and inducing severe 
overconfidence in Llama-Instruct. The effect is quantitatively 
amplified relative to the general-QA setting studied by Kadavath 
et al., with $\Delta\ece$ values reaching $+0.451$ in the worst case 
--- approximately three times the magnitude previously reported. This 
amplification is compatible with the intrinsic difficulty of the 
clinical task, which offers more room for confidence--accuracy 
mismatch than general-knowledge questions do.

Post-hoc isotonic recalibration (F6) provides a partial remedy: it 
brings eleven of the twelve configurations below the calibration 
threshold, robustly across five-fold cross-validation. It cannot, 
however, restore reasoning ability to models that do not classify. The 
practical conclusion is that isotonic recalibration should be part of 
any pre-deployment audit of LLM-based clinical prediction pipelines, 
but that recalibration cannot substitute for the choice of a model 
that produces diverse and clinically informed predictions in the first 
place.

% ----------------------------------------------------------------------------
\section{Contributions}
\label{sec:conc-contributions}

The scientific contributions of this thesis fall into three groups.

First, the study provides the first systematic calibration audit of 
open-source LLMs applied to clinical trial matching. The audit follows 
the reporting recommendations of TRIPOD+AI~\cite{collins2024tripod} and 
covers twelve configurations organized in a two-by-two experimental 
matrix (generalist versus medical, Base versus Instruct). Every 
prediction and every numerical result is archived in a public 
repository to support independent replication (Appendix~\ref{app:code}).

Second, the study extends the RLHF-degrades-calibration finding of 
Kadavath et al.~\cite{kadavath2022know} to the clinical domain, with a 
pre-registered permutation test replicated on two independent model 
families under two independent prompt designs. To our knowledge, this 
is the first replication of this effect outside general-purpose 
question answering, and the first quantification of its amplification 
in a high-stakes clinical task.

Third, the study documents three failure modes of open-source medical 
LLMs that had not been previously reported in a systematic form. The 
enriched-prompt paradox (F3) shows that expert prompt engineering can 
degrade calibration. Mode collapse (F7, F8) reveals that a substantial 
fraction of open-source medical models operate as biased constant 
predictors, and that this behavior persists under constrained decoding. 
Complete failure on discriminative cases (F9) demonstrates that global 
accuracy metrics can mask total task failure on the cases that carry 
the most clinical information. These three findings jointly motivate 
the composite utility score introduced in 
Section~\ref{subsec:met-composite}, which combines calibration, 
discrimination, and prediction diversity into a single rank-ordering 
that resists majority-class shortcuts.

The methodological byproducts of the study --- the pre-registered 
protocol, the double-prompt design, the log-probability scoring 
pipeline, and the composite utility score --- are intended as 
reusable elements for future calibration audits of clinical LLMs.

% ----------------------------------------------------------------------------
\section{Limitations}
\label{sec:conc-limitations}

Three categories of limitation qualify the conclusions of this work: 
the reliance on a single benchmark, the constrained model scale, and 
the finite prompt space explored.

\subsection{External Validation}
\label{subsec:conc-limitations-external}

The present study relies on a single benchmark, 
TrialGPT-Criterion-Annotations~\cite{jin2024trialgpt}, chosen because 
it is the only public corpus offering per-criterion multi-level expert 
annotations with an explicit inclusion/exclusion distinction --- 
properties that are essential for fine-grained calibration analysis. 
External validation on independent datasets was not carried out, for 
three complementary reasons. First, existing alternatives are not 
directly comparable: TREC Clinical Trials 2021--2022 and SIGIR 2016 
operate at the trial level rather than the criterion level and use 
only three ordinal relevance grades, which precludes direct 
replication of the calibration protocol used here. Adapting the 
hypotheses to these datasets would not constitute a validation but a 
distinct study with different endpoints. Second, the mode collapse 
phenomenon documented in Section~\ref{sec:res-mode-collapse} appears 
intrinsically linked to the availability of an evasion label in the 
label space; datasets that do not include such a label cannot be used 
to test whether this behavior generalizes. Third, prospective 
validation on real clinical data would require ethical approval, 
hospital partnership, and integration into a live recruitment 
workflow --- constraints that situate external validation as a 
research project in its own right, complementary to the present work 
rather than a subordinate step. External validation nevertheless 
remains an important direction for future work.

\subsection{Model Scale}
\label{subsec:conc-limitations-scale}

The models evaluated in this study are limited to 7--9 billion 
parameters, which is the size range that fits into a free-tier T4 GPU 
with 4-bit quantization. Larger medical LLMs, such as Meditron-70B or 
Med42-v2-70B, may exhibit qualitatively different calibration and 
failure-mode patterns. In particular, larger models may be more 
resistant to mode collapse, which would attenuate but not eliminate 
Findings F7 and F8. The extrapolation of the present results to larger 
scales is a natural direction for future work, and the methodology 
proposed here is directly transferable to 70B-scale configurations 
provided that the computational budget is available.

\subsection{Prompt Space}
\label{subsec:conc-limitations-prompts}

Only two prompt variants were evaluated in the present study: a 
minimal Prompt~A and an enriched Prompt~B. The space of possible 
prompts is much larger, encompassing chain-of-thought designs, 
role-based framings, and few-shot examples selected through active 
learning. The enriched-prompt paradox (F3) documented here is thus 
specific to the two prompts tested; it may or may not generalize to 
other prompt families. A systematic study of the calibration effect 
of prompt design is a promising direction for follow-up work.

% ----------------------------------------------------------------------------
\section{Future Work}
\label{sec:conc-future}

Four research directions extend naturally from the present study, in 
order of feasibility.

\subsection{External Validation on Independent Benchmarks}
\label{subsec:conc-external}

The most immediate extension is to validate the findings on 
independent datasets. TREC Clinical Trials 2021--2022, despite its 
coarser granularity, would provide a first replication of Hypothesis 
H1 in a different setting. MIMIC-derived cohorts and n2c2 datasets, 
which contain more heterogeneous patient descriptions, would test 
whether mode collapse persists across styles of clinical documentation. 
Ultimately, prospective validation on live recruitment workflows would 
close the gap between benchmark performance and clinical utility.

\subsection{Multi-Agent Adversarial Debate for Calibration}
\label{subsec:conc-debate}

The failure of individual models on discriminative cases motivates 
investigating whether ensembles of models can outperform any single 
one. Multi-agent debate architectures, in which several LLMs exchange 
arguments before converging on a joint prediction 
\cite{ouyang2022training}, are a particularly interesting candidate: 
models with distinct RLHF biases and distinct evasion channels may 
mutually neutralize each other's overconfidence through argumentative 
confrontation. To our knowledge, this paradigm has not been evaluated 
for clinical trial matching, and the calibration properties of the 
resulting consensus scores are unknown. The 6-model consensus analysis 
reported in Section~\ref{subsec:res-consensus}, which shows a 75.5\% 
accuracy on the 163 pairs of full agreement, suggests that this 
direction is promising.

\subsection{Investigating the Origin of Mode Collapse}
\label{subsec:conc-collapse-origin}

The mode collapse documented in Findings F7 and F8 raises a 
mechanistic question: is the collapse caused by safety-oriented 
alignment procedures that push models toward evasive answers, by 
domain shift between medical pretraining data and clinical trial 
criteria, or by a combination of both? Investigating this question 
would require access to the training corpora and reward models of the 
collapsed configurations, and possibly interpretability analyses of 
the underlying representations. A systematic investigation could 
inform the design of future medical LLMs so as to avoid the specific 
collapse patterns documented here.

\subsection{Extension to Larger and Multilingual Models}
\label{subsec:conc-scale-future}

Extending the analysis to 70-billion-parameter models is a 
straightforward but resource-intensive direction. Meditron-70B and 
Med42-v2-70B are the natural targets; both require GPU resources 
beyond the free-tier compute budget used here. Multilingual evaluation 
is a second direction: models such as MMed-Llama-3 and Apollo have 
been developed to support clinical text in languages other than 
English, and the calibration properties of these multilingual models 
have not been systematically audited on any language.

% ----------------------------------------------------------------------------
\section{Closing Remarks}
\label{sec:conc-closing}

The systematic use of large language models in clinical prediction 
workflows is likely to expand rapidly in the coming years, driven by 
the maturation of open-source alternatives, the pressure to accelerate 
clinical trial recruitment, and the availability of increasingly 
capable models. This trajectory makes the auditing of these models a 
practical necessity rather than an academic exercise. The present 
thesis contributes to this auditing effort in three specific ways: it 
proposes a reusable methodology, it documents three previously 
unreported failure modes in open-source medical LLMs, and it archives 
its data and code for independent verification.

The broader lesson of this study is that global performance metrics 
--- accuracy, F1, and even ECE taken in isolation --- are insufficient 
to characterize the clinical usability of a language model. A 
responsible evaluation protocol must combine calibration with 
diversity, must stratify performance by case difficulty, and must 
remain vigilant against the possibility of mode collapse. The 
composite utility score proposed in this thesis is one such combined 
protocol; developing more refined alternatives is an important 
direction for the calibration and clinical AI communities alike.

None of the twelve open-source configurations audited here provides a 
fully satisfactory basis for autonomous clinical trial matching. The 
findings do, however, identify Mistral-Instruct with a minimal prompt 
as the most promising candidate for further study, and identify 
post-hoc isotonic recalibration as a practical remedy for the 
miscalibration that afflicts all configurations. Whether these two 
elements are jointly sufficient for a deployment scenario --- with 
mandatory human oversight, ongoing calibration monitoring, and 
appropriate protocols for handling low-confidence cases --- is a 
question that the present thesis leaves to prospective clinical 
validation. Answering it is the natural next step of this line of 
research.