% ============================================================================
%   APPENDIX D — CODE AND DATA AVAILABILITY
% ============================================================================

\chapter{Code and Data Availability}
\label{app:code}

This appendix documents the archival of the code, prompts, prediction 
files, and analysis outputs produced during this study, in accordance 
with the reproducibility commitments described in 
Section~\ref{sec:met-reproducibility}.

% ----------------------------------------------------------------------------
\section{Repository}
\label{sec:app-repo}

The complete project is archived in a public GitHub repository:

\begin{center}
    \url{https://github.com/gyxcit/llm-clinical-trial-calibration}
\end{center}

The repository is organized into three top-level directories, each 
serving a distinct purpose:

\begin{itemize}
    \item \texttt{thesis/} contains the \LaTeX{} sources of this 
    manuscript, including the main file, all chapters, appendices, 
    the bibliography, and the figures embedded in the document.
    \item \texttt{code/} contains the Kaggle notebook used for the 
    twelve model runs and the Python scripts that perform the 
    downstream analyses (calibration audit, hypothesis tests, 
    recalibration, conformal prediction, mode collapse analysis, 
    composite utility scoring, the signal-to-prior mechanistic 
    analysis, the out-of-fold re-thresholding, and the TREC external 
    replication), together with the figure-generation script that 
    produces the diagrams in both French and English.
    \item \texttt{data/} contains the raw prediction files (twelve 
    Parquet files, one per configuration) and the CSV outputs of 
    every downstream analysis reported in Chapter~\ref{chap:results}.
\end{itemize}

The repository is released under an MIT license 
(\texttt{LICENSE}~file at the root), which permits redistribution 
and derivative use for research purposes.

% ----------------------------------------------------------------------------
\section{Data Availability}
\label{sec:app-data}

The primary benchmark used in this study, 
TrialGPT-Criterion-Annotations, is publicly available under a CC BY 
4.0 license as a supplementary release of Jin et 
al.~\cite{jin2024trialgpt}. It can be downloaded directly from the 
TrialGPT repository:

\begin{center}
    \url{https://github.com/ncbi-nlp/TrialGPT}
\end{center}

The external validation corpus is derived from the TREC Clinical 
Trials 2021--2022 tracks~\cite{roberts2021trec, roberts2022trec}, whose 
topics and relevance judgements are publicly available from the 
National Institute of Standards and Technology (NIST). The scripts that 
sample the topics, decompose trials into patient--criterion pairs, and 
score them with the log-probability pipeline are included in the 
\texttt{code/} directory, and the resulting TREC prediction files are 
archived alongside the primary prediction files in \texttt{data/}.

The twelve prediction files produced during this study are stored in 
the \texttt{data/} subdirectory of the project repository, one 
Parquet file per configuration. Each file contains, for the $1{,}015$ 
patient--criterion pairs, the following columns: the annotation 
identifier, the criterion type, the predicted label, the associated 
confidence, the expert consensus label, the GPT-4 label, and the 
full softmax distribution over the six candidate labels 
(\texttt{p\_included}, \texttt{p\_not\_included}, \texttt{p\_excluded}, 
\texttt{p\_not\_excluded}, \texttt{p\_not\_enough\_information}, 
\texttt{p\_not\_applicable}). All downstream analysis outputs --- 
calibration metrics, permutation-test results, recalibration 
comparisons, conformal prediction sets, mode collapse statistics, 
composite utility scores, the signal-to-prior diagnostic, the 
out-of-fold re-thresholding gains, and the TREC replication --- are 
also archived as CSV files in the same directory. Each numerical value 
reported in Chapter~\ref{chap:results} and in Appendices~\ref{app:tables} 
and~\ref{app:mode-collapse} can be traced to a specific row of one of 
these files.

% ----------------------------------------------------------------------------
\section{Reproducing the Experiments}
\label{sec:app-reproduce}

Independent reproduction of the pipeline follows five steps:

\begin{enumerate}
    \item Clone the repository and install the Python dependencies.
    \item Download the TrialGPT-Criterion-Annotations dataset from 
    the URL above and place it in the input directory expected by 
    the notebook.
    \item Obtain access to the six selected models from the 
    HuggingFace Hub. Llama-3.1 requires acceptance of Meta's 
    community license, obtained through the HuggingFace interface.
    \item Execute the Kaggle notebook (or its local equivalent) to 
    produce the twelve prediction files. Random seeds are fixed at 
    $42$ throughout the pipeline to ensure exact reproducibility.
    \item Execute the analysis cells in the order documented in the 
    notebook to reproduce every table and figure of 
    Chapter~\ref{chap:results}.
\end{enumerate}

All data splits used by the downstream analyses are deterministic and 
seeded, so that any reviewer reproduces the exact partitions used here. 
The post-hoc recalibration (Section~\ref{subsec:met-recalibration}) uses 
a $70/30$ split stratified by criterion type with \texttt{random\_state} 
$=42$. The stability analyses --- the five-fold cross-validation of 
isotonic recalibration and the out-of-fold cross-validation that fits 
the re-thresholding decision boundary 
(Section~\ref{subsec:met-rethreshold}) --- use a 
\texttt{KFold} splitter with $k=5$, \texttt{shuffle=True}, and the same 
seed $42$. The bootstrap confidence intervals throughout the thesis draw 
$1{,}000$ resamples under the same seed. Because the splitter, the 
number of folds, and the seed are all fixed and recorded in the 
notebook, the fold assignments, the fitted thresholds, and the resulting 
confidence intervals are bit-exact reproducible across independent runs.

% ----------------------------------------------------------------------------
\section{Computational Requirements}
\label{sec:app-compute}

The twelve model runs require a GPU with at least $16$~GB of memory 
(a Tesla~T4 or equivalent) and approximately ten to fifteen hours of 
GPU time in total. The downstream analyses run on CPU and complete 
in a few minutes on standard hardware. All experiments reported in 
this thesis were performed on the Kaggle free-tier compute 
environment, which is publicly accessible and imposes no barriers to 
independent replication.

% ----------------------------------------------------------------------------
\section{Software Versions}
\label{sec:app-versions}

The pipeline was tested with the following software versions:

\begin{itemize}
    \item Python 3.12.13
    \item PyTorch 2.10.0 with CUDA 12.8
    \item \texttt{transformers} 5.0.0
    \item \texttt{scikit-learn} 1.6.1
    \item \texttt{numpy} 2.4.6, \texttt{pandas} 2.3.3, \texttt{scipy} 1.16.3
    \item \texttt{matplotlib} 3.10.0
    \item \texttt{bitsandbytes} (for 4-bit NF4 quantization; version compatible with the PyTorch build listed above)
\end{itemize}

Version pinning is documented in the notebook itself, in the cell 
that installs the dependencies.