% ============================================================================
%   APPENDIX F — DETAILED MODE COLLAPSE ANALYSIS
% ============================================================================

\chapter{Detailed Mode Collapse Analysis}
\label{app:mode-collapse}

This appendix documents in full detail the mode collapse phenomenon 
introduced in Section~\ref{sec:res-mode-collapse}. Five analyses are 
reported: the complete distribution of predictions across the six 
labels for every configuration; the outcomes of the constrained-decoding 
simulation that established the persistence of the collapse under a 
change of candidate set (Finding F8); the mechanistic analysis that 
reinterprets the collapse as a decoding artefact through the 
signal-to-prior ratio; the extended comparison between dominant-class 
and truly-discriminative cases (Finding F9); and a quantitative link 
between prediction diversity and reasoning ability on discriminative 
subsets.

% ----------------------------------------------------------------------------
\section{Complete Prediction Distributions}
\label{sec:app-distributions}

Table~\ref{tab:app-distributions} reports the proportion of predictions 
assigned to each of the six labels for every configuration. The values 
are computed on the full $1{,}015$ patient--criterion pairs. Rows sum to 
$100\%$ up to rounding.

\begin{table}[htbp]
    \centering
    \caption{Complete prediction distributions across the six labels 
        for the twelve configurations. \emph{Incl.}: included. \emph{NIncl.}: 
        not included. \emph{Excl.}: excluded. \emph{NExcl.}: not excluded. 
        \emph{NEI}: not enough information. \emph{NA}: not applicable.}
    \label{tab:app-distributions}
    \small
    \begin{tabular}{@{}lcccccc@{}}
        \toprule
        \textbf{Configuration} & 
        \textbf{Incl.} & \textbf{NIncl.} & 
        \textbf{Excl.} & \textbf{NExcl.} & 
        \textbf{NEI} & \textbf{NA} \\
        \midrule
        Mistral-Base A     &  5.4\% &  1.0\% & 12.5\% & 47.2\% & 33.3\% &  0.6\% \\
        Mistral-Base B     &  9.7\% &  4.9\% &  5.3\% & 45.1\% & 34.2\% &  0.8\% \\
        Mistral-Instruct A &  8.1\% &  3.6\% &  4.6\% & 27.4\% & 47.4\% &  8.9\% \\
        Mistral-Instruct B & 14.8\% &  6.4\% & 32.5\% & 20.3\% & 20.8\% &  5.2\% \\
        Llama-Base A       &  4.3\% &  1.2\% & 45.4\% & 22.9\% & 26.2\% &  0.0\% \\
        Llama-Base B       &  0.8\% &  0.5\% &  0.5\% &  0.6\% & 97.2\% &  0.4\% \\
        Llama-Instruct A   &  0.3\% &  0.4\% &  0.6\% &  0.9\% & 97.8\% &  0.0\% \\
        Llama-Instruct B   &  0.1\% &  0.0\% &  0.1\% &  0.1\% & 99.7\% &  0.0\% \\
        Meditron A         &  6.1\% &  2.1\% & 22.4\% & 64.8\% &  4.6\% &  0.0\% \\
        Meditron B         & 10.4\% &  4.8\% & 21.3\% & 59.3\% &  4.2\% &  0.0\% \\
        BioMistral A       &  0.5\% &  0.4\% &  0.5\% &  0.4\% & 98.2\% &  0.0\% \\
        BioMistral B       &  0.0\% &  0.0\% &  0.0\% &  0.0\% & 100.0\% & 0.0\% \\
        \bottomrule
    \end{tabular}
\end{table}

Five configurations concentrate more than $97\%$ of their predictions 
on the same label (\emph{not enough information}): both BioMistral 
variants, both Llama-Instruct variants, and Llama-Base with Prompt~B. 
Two configurations exhibit partial collapse (Meditron~A and B, with 
\emph{not excluded} accounting for $59.3\%$--$64.8\%$ of predictions). 
The five remaining configurations distribute their predictions more 
evenly across labels, though none exceeds $50\%$ on any single class.

% ----------------------------------------------------------------------------
\section{Constrained-Decoding Simulation (Finding F8)}
\label{sec:app-forced}

To distinguish whether mode collapse reflects a genuine absence of 
clinical reasoning or an artifact of the six-label vocabulary, we 
conducted a controlled simulation. For each collapsed configuration, 
the softmax normalization (Equation~\ref{eq:softmax}) was restricted 
to the two decisive labels of each criterion type: 
$\{$\emph{included}, \emph{not included}$\}$ for inclusion criteria, 
and $\{$\emph{excluded}, \emph{not excluded}$\}$ for exclusion criteria. 
The evasion labels were removed from the candidate set, forcing the 
model to commit to a decisive answer.

Table~\ref{tab:app-forced} reports the prediction distribution under 
this constrained regime, restricted to the five collapsed 
configurations of Section~\ref{sec:app-distributions}. Under 
constrained decoding, all five configurations concentrate their 
predictions on the majority label of the restricted candidate set 
(\emph{not excluded} for exclusion criteria, \emph{included} for 
inclusion criteria), with proportions similar to the original 
collapse.

\begin{table}[htbp]
    \centering
    \caption{Prediction distribution under constrained decoding 
        (evasion labels removed from candidate set). All five 
        collapsed configurations migrate their collapse to the 
        majority label of the restricted set. Accuracy is computed 
        against the expert consensus label.}
    \label{tab:app-forced}
    \small
    \begin{tabular}{@{}lccc@{}}
        \toprule
        \textbf{Configuration} & 
        \textbf{Modal label} & 
        \textbf{Usage rate} & 
        \textbf{Accuracy} \\
        \midrule
        BioMistral A      & not excluded / included & 74.2\% & 0.292 \\
        BioMistral B      & not excluded / included & 76.8\% & 0.294 \\
        Llama-Base B      & not excluded / included & 73.5\% & 0.311 \\
        Llama-Instruct A  & not excluded / included & 78.4\% & 0.298 \\
        Llama-Instruct B  & not excluded / included & 81.1\% & 0.292 \\
        \bottomrule
    \end{tabular}
\end{table}

The constrained-decoding accuracy remains near the base rate of the 
majority class ($47.7\%$ if all exclusion criteria are predicted 
\emph{not excluded}, weighted down by the inclusion cases). The 
collapse thus persists under a change of candidate set: forcing the 
model to choose between two decisive labels does not restore a diverse 
$\arg\max$, it simply redirects the concentration of predictions from an 
evasion label to the numerically dominant decisive label. This 
experiment grounds Finding~F8 --- the collapse is a property of the 
model's decision, not of the evasion vocabulary. It does \emph{not}, 
however, establish that the model lacks discriminative knowledge: the 
constrained decoding still takes the $\arg\max$, which the next section 
shows to be exactly the operation that masks a signal present in the 
underlying probabilities. The distinction between a collapsed $\arg\max$ 
and an absent signal is the subject of 
Section~\ref{sec:app-mechanism}.

% ----------------------------------------------------------------------------
\section{The Signal-to-Prior Mechanism}
\label{sec:app-mechanism}

Section~\ref{subsec:res-collapse-mechanism} showed that mode collapse is 
governed by the ratio between the amplitude with which a model varies the 
positive-label log-probability across inputs (the signal, 
$\sigma_{\text{pos}}$) and the systematic offset by which competing 
labels are preferred (the response prior, $\text{bias}_{\text{ext}}$). 
This appendix reports the underlying quantities for all twelve 
configurations, together with the ordinal separation (Cliff's $\delta$) 
that establishes the presence of a discriminative signal even under a 
collapsed $\arg\max$.

Table~\ref{tab:app-mechanism} reports, per configuration, the signal 
amplitude, the response prior, their ratio $r$, and Cliff's $\delta$ 
between the positive-label probability on gold-positive versus 
gold-negative inclusion cases. The configurations are sorted by $r$.

\begin{table}[htbp]
    \centering
    \caption{Signal amplitude ($\sigma_{\text{pos}}$), response prior 
        ($\text{bias}_{\text{ext}}$, in nats), their ratio 
        $r = \sigma_{\text{pos}} / |\text{bias}_{\text{ext}}|$, and the 
        ordinal separation (Cliff's $\delta$) on inclusion criteria. A 
        high $\delta$ with a low $r$ is the signature of a live signal 
        masked by the $\arg\max$; a negative $\delta$ (Meditron) indicates 
        a corrupted signal. Sorted by $r$.}
    \label{tab:app-mechanism}
    \small
    \begin{tabular}{@{}lcccc@{}}
        \toprule
        \textbf{Configuration} & 
        \boldmath$\sigma_{\text{pos}}$ & 
        \boldmath$\text{bias}_{\text{ext}}$ & 
        \boldmath$r$ & 
        \textbf{Cliff's $\delta$} \\
        \midrule
        Mistral-Base B     & 0.61 & $+0.32$ & 1.92 & $+0.539$ \\
        Mistral-Instruct A & 5.85 & $+3.21$ & 1.82 & $+0.741$ \\
        Mistral-Instruct B & 4.47 & $+2.99$ & 1.49 & $+0.665$ \\
        Llama-Base A       & 0.21 & $+0.16$ & 1.32 & $+0.293$ \\
        Mistral-Base A     & 0.28 & $-0.38$ & 0.75 & $+0.464$ \\
        Llama-Instruct B   & 2.61 & $+5.69$ & 0.46 & $+0.733$ \\
        Llama-Instruct A   & 1.42 & $+4.04$ & 0.35 & $+0.823$ \\
        Llama-Base B       & 0.39 & $+1.34$ & 0.29 & $+0.353$ \\
        BioMistral B       & 0.33 & $+1.29$ & 0.25 & $+0.499$ \\
        BioMistral A       & 0.19 & $+1.35$ & 0.14 & $+0.103$ \\
        Meditron A         & 0.19 & $+2.05$ & 0.09 & $-0.138$ \\
        Meditron B         & 0.16 & $+2.47$ & 0.07 & $+0.110$ \\
        \bottomrule
    \end{tabular}
\end{table}

Two facts stand out from Table~\ref{tab:app-mechanism}. First, the 
collapsed Llama-Instruct and BioMistral configurations combine a small 
signal-to-prior ratio with a positive Cliff's $\delta$: the 
discriminative ordering of the positive-label probability is preserved 
even though the $\arg\max$ never varies. The discriminative information 
is present --- it is simply not expressed at the decision step. Second, 
the two Meditron configurations have the smallest ratios of the entire 
study ($0.07$--$0.09$), and their ordinal separation collapses toward 
zero: Meditron~A is outright negative ($\delta = -0.138$), ranking cases 
in the wrong order, and Meditron~B is only marginally positive 
($\delta = +0.110$), well below the separation of every recoverable 
configuration. Neither carries a signal that a decision rule could 
exploit. This is the empirical foundation of the 
collapse-versus-corruption distinction developed in 
Section~\ref{subsec:res-rethreshold-diag}: a low ratio combined with a 
strong positive $\delta$ marks recoverable expressive collapse, whereas a 
low ratio combined with a near-zero or negative $\delta$ marks 
irrecoverable corruption. It is the discriminative separation, captured 
in the main text by the AUC, that distinguishes the two regimes --- not 
the ratio or the predicted-label distribution alone.

The Spearman correlation between $r$ and the $\arg\max$ macro-F1 on 
inclusion criteria is $\rho = 0.866$ ($p = 0.00027$, $n = 12$): the ratio 
predicts, almost monotonically, whether a configuration classifies or 
collapses. The separation is clean at $r \approx 0.6$ --- every 
configuration above it classifies, every configuration below it collapses 
--- which is the threshold used to delimit the healthy regime in 
Section~\ref{subsec:res-rethreshold-diag}.

% ----------------------------------------------------------------------------
\section{Dominant vs.\ Discriminative Cases (Finding F9)}
\label{sec:app-discriminative}

Table~\ref{tab:app-discriminative} extends Table~\ref{tab:res-non-trivial} 
of Section~\ref{sec:res-non-trivial} with the full breakdown across all 
twelve configurations, sorted by discriminative accuracy. The 
dominant-class subset consists of the $777$ pairs whose expert label 
is either \emph{not excluded} or \emph{not enough information}. The 
discriminative subset consists of the $238$ remaining pairs whose 
expert label is one of \emph{included}, \emph{excluded}, 
\emph{not included}, or \emph{not applicable}.

\begin{table}[htbp]
    \centering
    \caption{Extended dominant-vs.-discriminative comparison. The 
        discriminative accuracy on the $238$ pairs is the metric that 
        best reflects clinical reasoning ability. Positive drops 
        indicate that the configuration performs better on the 
        discriminative subset than on the dominant subset.}
    \label{tab:app-discriminative}
    \small
    \begin{tabular}{@{}lccc@{}}
        \toprule
        \textbf{Configuration} & 
        \textbf{Dominant acc.} & 
        \textbf{Discriminative acc.} & 
        \boldmath$\Delta$ \\
        \midrule
        Mistral-Instruct A & 0.183 & 0.719 & $+0.536$ \\
        Mistral-Base A     & 0.462 & 0.613 & $+0.151$ \\
        Mistral-Instruct B & 0.474 & 0.500 & $+0.026$ \\
        Mistral-Base B     & 0.638 & 0.416 & $-0.222$ \\
        Llama-Base A       & 0.219 & 0.265 & $+0.046$ \\
        Llama-Instruct A   & 0.373 & 0.021 & $-0.352$ \\
        Llama-Base B       & 0.398 & 0.013 & $-0.385$ \\
        Llama-Instruct B   & 0.383 & 0.013 & $-0.370$ \\
        BioMistral A       & 0.386 & 0.004 & $-0.382$ \\
        BioMistral B       & 0.377 & 0.000 & $-0.377$ \\
        Meditron A         & 0.846 & 0.000 & $-0.846$ \\
        Meditron B         & 0.822 & 0.000 & $-0.822$ \\
        \bottomrule
    \end{tabular}
\end{table}

Meditron exhibits the widest gap between the two subsets: $84.6\%$ 
accuracy on the $777$ dominant-class pairs but $0.0\%$ on the $238$ 
discriminative pairs. BioMistral shows a similar reversal. These 
patterns are the empirical basis of Finding~F9. Only 
Mistral-Instruct~A reverses the direction and performs substantially 
better on the discriminative subset ($71.9\%$) than on the dominant 
subset ($18.3\%$), suggesting that this configuration is the only 
one whose predictions reflect actual clinical judgment rather than 
majority-class shortcuts.

% ----------------------------------------------------------------------------
\section{Diversity Predicts Discriminative Accuracy}
\label{sec:app-diversity-accuracy}

The prediction diversity metric (normalized Shannon entropy, 
Equation~\ref{eq:diversity}) tracks discriminative accuracy closely. 
Table~\ref{tab:app-diversity-accuracy} reports the two metrics side by 
side, sorted by prediction diversity. The Pearson correlation between 
prediction diversity and discriminative accuracy across the twelve 
configurations is $r = 0.89$ ($p < 0.001$, $n = 12$).

\begin{table}[htbp]
    \centering
    \caption{Prediction diversity and discriminative accuracy, sorted 
        by diversity. Configurations with near-zero diversity produce 
        near-zero accuracy on the $238$ discriminative pairs.}
    \label{tab:app-diversity-accuracy}
    \small
    \begin{tabular}{@{}lcc@{}}
        \toprule
        \textbf{Configuration} & 
        \textbf{Diversity} & 
        \textbf{Discriminative acc.} \\
        \midrule
        Mistral-Instruct B & 0.842 & 0.500 \\
        Mistral-Instruct A & 0.812 & 0.719 \\
        Mistral-Base A     & 0.686 & 0.613 \\
        Mistral-Base B     & 0.577 & 0.416 \\
        Llama-Base A       & 0.525 & 0.265 \\
        Meditron B         & 0.377 & 0.000 \\
        Meditron A         & 0.362 & 0.000 \\
        Llama-Base B       & 0.077 & 0.013 \\
        Llama-Instruct A   & 0.066 & 0.021 \\
        BioMistral A       & 0.055 & 0.004 \\
        Llama-Instruct B   & 0.011 & 0.013 \\
        BioMistral B       & 0.000 & 0.000 \\
        \bottomrule
    \end{tabular}
\end{table}

This correlation supports the design choice of the composite utility 
score (Section~\ref{subsec:met-composite}), in which prediction 
diversity enters multiplicatively. A configuration with high 
calibration but zero diversity cannot produce discriminative 
predictions, and its clinical utility is correctly captured as zero 
by the composite score. Conversely, a configuration with modest 
calibration but high diversity retains the potential for discriminative 
reasoning that can be surfaced through recalibration or ensemble 
methods.

The magnitude of this correlation ($r^2 = 0.79$) indicates that 
prediction diversity alone accounts for nearly $80\%$ of the variance 
in discriminative accuracy \emph{under the default $\arg\max$ rule}. 
This near-mechanical relationship makes prediction entropy a cheap 
early-warning signal for collapse, detectable from the prediction 
distribution alone without any access to expert labels: monitoring the 
entropy over a moving window of recent inputs flags a collapsed 
configuration before any ground truth is available. It is important, 
however, not to over-read the correlation as a verdict of clinical 
unsuitability. Diversity measures whether the $\arg\max$ expresses a 
signal, not whether a signal exists: as Section~\ref{sec:app-mechanism} 
and the re-thresholding results (Section~\ref{sec:res-rethreshold}) show, 
several near-zero-diversity configurations carry a strong latent signal 
that a non-$\arg\max$ decision rule recovers. Low diversity is therefore 
a reliable flag that a model is unusable \emph{as deployed}, and a 
prompt to apply the signal/prior diagnostic --- not a conclusion that 
the model is beyond repair.