% ============================================================================
%   APPENDIX E — GENERATIVE AI DECLARATION (Aivancity mandatory)
% ============================================================================

\chapter{Generative AI Usage Declaration}
\label{app:ai-declaration}

% Aivancity guidelines require a transparent declaration of any use of
% generative AI tools during the preparation of this work.

% [TODO — 1-2 pages, be precise and honest]

\section*{Tools Used}

During the preparation of this thesis, the following generative AI tool was
used as an assistant:

\begin{itemize}
    \item \textbf{Claude} (Anthropic), primarily the Sonnet and Opus series
        model families, accessed via the \texttt{claude.ai} interface.
\end{itemize}

\section*{Purpose of Use}

Generative AI was used for:

\begin{itemize}
    \item \textbf{Ideation and design discussions}: brainstorming experimental
        designs, discussing methodological trade-offs, and refining the
        research question.
    \item \textbf{Code assistance}: writing and debugging Python code for the
        evaluation pipeline, prompt engineering iterations, and statistical
        analyses.
    \item \textbf{Writing assistance}: refining the phrasing of prompts,
        drafting progress reports, and revising English formulations of
        methodological descriptions.
    \item \textbf{Bibliographic exploration}: identifying candidate references
        and papers relevant to the state of the art.
\end{itemize}

\section*{Purpose of Non-Use}

Generative AI was \emph{not} used for:

\begin{itemize}
    \item Generating experimental data or results.
    \item Fabricating statistical outcomes or numerical values.
    \item Producing scientific claims not backed by our own experiments.
    \item Writing the final thesis text verbatim (the author remains
        responsible for all formulations).
\end{itemize}

\section*{Verification and Author Responsibility}

All AI-generated content used in this work was:

\begin{itemize}
    \item Reviewed, verified, and adapted by the author before inclusion.
    \item Cross-checked against primary sources for any factual claim.
    \item Understood by the author, who takes full responsibility for the
        scientific and methodological choices reflected in this thesis.
\end{itemize}

The scientific reasoning, experimental design, execution, and interpretation
of results are the sole responsibility of the author.
