% ============================================================================
%   ACKNOWLEDGEMENTS
% ============================================================================

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\markboth{Acknowledgements}{}

I would like to express my sincere gratitude to Dr.\ Anuradha Kar, my 
academic supervisor at aivancity, for her patient guidance throughout 
this final year project. The regular exchanges with her helped shape 
the research question from an initial broad interest in medical AI 
into the specific, testable framework presented in this thesis, and 
her insistence on methodological rigor was a decisive influence on the 
pre-registered design adopted here.

I am grateful to the faculty and staff of aivancity for providing a 
stimulating academic environment during the two years of the Programme 
Grande École, and for supporting the choice of a purely academic 
research project as an alternative to the traditional industry 
internship. The freedom this arrangement offered was a rare privilege 
and made possible the depth of analysis reported in this work.

I acknowledge the authors of the TrialGPT-Criterion-Annotations 
dataset~\cite{jin2024trialgpt} for releasing their annotations under 
an open license, without which this thesis would not exist. I also 
thank the maintainers of the open-source model ecosystem --- Mistral 
AI, Meta, EPFL for Meditron, and Yanis Labrak and colleagues for 
BioMistral --- for making their models freely available for research 
audits of this kind.

I thank Kaggle for the free-tier GPU access that supported the 
experimental phase of this project. The twelve model runs would not 
have been feasible without the generous computational resources 
Kaggle offers to student researchers.

Finally, I would like to thank my family and close friends for their 
patience and encouragement throughout these months of intensive 
writing. Their support made the completion of this project 
substantially easier than it would otherwise have been.