% ============================================================================
%   ABSTRACT (English) — Aivancity: 250-300 words, ≥5 keywords, self-contained
% ============================================================================

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\markboth{Abstract}{}

Patient matching to clinical trials is a bottleneck in translational 
medicine, and recent large language model (LLM) systems such as TrialGPT 
aim to automate the assessment of patient eligibility against inclusion 
and exclusion criteria. Beyond raw accuracy, however, their probabilistic 
reliability has not been systematically audited: a recent scoping review 
of twenty-four LLM-based matching systems found none that report 
calibration metrics. Yet an overconfident model that misclassifies with 
high confidence is clinically worse than an accurate but honestly 
uncertain one.

We conducted the first systematic calibration audit of open-source LLMs 
in this domain: twelve configurations (six models $\times$ two prompts) 
evaluated on the 1{,}015 patient--criterion pairs of 
TrialGPT-Criterion-Annotations, using direct log-probability scoring and 
four pre-registered hypotheses with Bonferroni-corrected permutation 
tests.

All twelve configurations exceed the clinical threshold of $0.05$ 
(Expected Calibration Error from $0.124$ to $0.602$), and RLHF 
significantly degrades calibration on both model families ($p < 0.003$). 
Five configurations exhibit mode collapse, and both medical LLMs achieve 
accuracy at or below $0.4\%$ on the $238$ discriminative cases. We then 
show this collapse to be largely a decoding artefact rather than a 
knowledge loss: the discriminative signal survives in the 
log-probabilities (area under the curve up to $0.91$) but is masked by a 
response prior. A single scalar, the signal-to-prior ratio, predicts 
collapse and explains why monotone recalibration cannot repair it while a 
simple re-thresholding rule restores up to $+0.79$ macro-F1. The analysis 
separates recoverable collapse from irrecoverable corruption, and both 
replicate on the independent TREC Clinical Trials corpus.

Aggregate metrics can thus mask complete failure on clinically decisive 
cases, and can equally mask a recoverable signal beneath an apparent 
failure. A responsible evaluation protocol must combine calibration, 
prediction diversity, stratified analysis by case difficulty, and a 
diagnostic distinguishing a collapsed decision rule from a corrupted 
signal.

\vspace{0.5cm}
\noindent\textbf{Keywords:} large language models, calibration, 
clinical trial matching, RLHF, medical AI, mode collapse, expected 
calibration error, re-thresholding.

\cleardoublepage